---
title: "Organiser Essentials" 
subtitle: "Inflation, Wages, and Cost of Living"
author: "Hamish Gamble"
affiliation: "National Research Centre" 
execute:    
  echo: false
  warning: false
  message: false
format:
  revealjs:
      embed-resources: true
      auto-play-media: true
      incremental: false
      slide-number: true
      fig-format: svg
      
  pptx:
    incremental: false
    slide-level: 2 
    fig-format: svg
    reference-doc: template.pptx
brand: _brand/_brand.yml
editor: visual 
---

```{r}
#| label: setup
library(tidyverse)
library(readabs)
library(reportabs)
library(directlabels)
library(scales)
library(glue)
library(gt)
library(geomtextpath)

cpi_quarterly_group <- read_absdata("cpi_quarterly_group", export_dir = "data", refresh = TRUE)

wpi_quarterly <- read_absdata("wpi_quarterly", export_dir = "data", refresh = TRUE)

cli_quarterly <- read_absdata("cli", export_dir = "data", refresh = TRUE)

plot_real_inflation_group <- function(cpi_category, since) {
  
  if (cpi_category == "Housing") {
    cpi <- cli_quarterly |> 
      filter(cpi_category == "Housing",
             data_type == "Index Numbers",
             household_type == "Employee households",
             date >= {{since}}) |> 
      mutate(value = value/value[date=={{since}}] - 1,
             indicator = "Cost of Living Index") |> 
      select(date, indicator, value)
  } else {
  cpi <- cpi_quarterly_group |> 
    filter(cpi_category == {{cpi_category}},
           unit == "Index Numbers",
           date >= {{since}}) |> 
    mutate(value = value/value[date=={{since}}] - 1,
           indicator = "Consumer Price Index") |> 
    select(date, indicator, value)
  }
  
  wpi <- wpi_quarterly |> 
    filter(data_type == "Quarterly Index", 
           date >= {{since}},
           state == "Australia",
           indicator == "Ordinary time hourly rates of pay excluding bonuses",
           sector == "Private and Public",
           industry == "Manufacturing") |> 
    mutate(value = value/value[date=={{since}}] -1,
           indicator = "Manufacturing Wages") |> 
    select(date, indicator, value)
  
  date_label <- label_date("%B %Y")(last(cpi$date))
  since_label <- label_date("%B %Y")(as_date(since))
  cpi_label <- label_percent(accuracy = 0.1)(last(cpi$value))
  wpi_label <- label_percent(accuracy = 0.1)(last(wpi$value))
  
  plot_data <- bind_rows(cpi, wpi)  |>
    group_by(indicator) |> 
    mutate(label = label_percent(accuracy = 0.1)(last(value))) |> 
    ungroup()
  
  inc_dec <- ifelse(last(cpi$value) > 0, "increased", "decreased")
  
  ggplot(plot_data, aes(x = date, y = value, col = indicator)) + 
    geom_line() +
    labs(x = NULL,
         y = NULL,
         title = glue("{cpi_category} {inc_dec} by {cpi_label} between {since_label} and {date_label}"),
         subtitle = glue("Manufacturing wages increased by {wpi_label} between {since_label} and {date_label}")) + 
    scale_x_date(breaks = plot_date_range(plot_data),
                 date_labels = "%b-%y") +
    scale_y_continuous(labels = label_percent()) +
    geom_dl(aes(label = label), method = "last.qp") +
    theme_nrc(legend = "bottom")
  
  
}


```

## Organiser Essentials

::: notes
This is the second session in the NRC Legal, Political and Economic Education series which we are developing. These sessions are designed to:

-   help build a culture of mindful militancy
-   have a shared understanding (as a union) on legal, political, and economic issues
-   provide consistency in training across delegates and organisers, to help with coaching and mentoring of delegates following DU4 and more broadly around knowledge and skill building.
-   provide some useful talking points in recruitment and organising conversations.
-   What comes next will be influenced by what you think you need/what you think will be useful
:::

-   These sessions are designed to help build a culture of mindful militancy.

-   Consistency in education across members, delegates, and organisers

-   Tell us what you need!

## Talking About Inflation

::: notes
Understanding Inflation is key to why:

-   Cost of living is not accurately reflected in the CPI

-   Limitations of tying wage increases to CPI
:::

-   Why do we need to understand inflation?

-   What is inflation, and what is it not?

-   Talking about productivity in the workplace

## Talking About Inflation

::: notes
So over the past 3 years, wages have actually grown faster than CPI. Job done, problem solved.
:::

```{r}

plot_real_inflation_group("All groups CPI", "2023-03-01")

```

{{< include sections/_section-02a-inflation.qmd >}}

## Inflation Impact - Essential Goods and Services

::: notes
:::

## Bread

```{r}



```

## Milk

```{r}

```

## Gas

```{r}

```

## Electricity

```{r}

```

## Property Rates

```{r}

```

## Rents

```{r}

```

## Water

```{r}

```

## Food and Non-alcoholic Beverages

```{r}

```

## Housing - Including Mortgages

```{r}



                                                                        

```

## Change in Price relative to Wages

```{r}


```

## So how much should we ask for?

```{r}


```

## Wages in Manufacturing

```{r}

```
