# Housing

```{r}
#|label: housing-setup
library(tidyverse)
library(readabs)
library(reportabs)
library(gt)
library(systemfonts)

knitr::opts_chunk$set(dev = "ragg_png")
# Required data for house price stuff:
# Total Value of Dwellings (Table 1)
mean_house_price <- read_abs("6432.0", tables = 1) |>
  filter(series == "Mean price of residential dwellings ;  Australia ;") |>
  select(date, mean_house_price = value)
# House price index
hpi <- read_csv(
  "https://stats.bis.org/api/v2/data/dataflow/BIS/WS_SPP/1.0/Q.AU.N.628?format=csv"
) |>
  select(date = TIME_PERIOD, house_price_index = OBS_VALUE) |>
  mutate(date = yq(date) + months(2))
# National accounts
national_accounts <- read_abs("5206.0", tables = c(1, 2, 20))
# Table 1 (real GDP per capita)
# Table 2 (real GDP)
# Table 20 (gross disposable income)
gdi_pc <- national_accounts |>
  filter(
    series %in%
      c(
        "GROSS DISPOSABLE INCOME ;",
        "GDP per capita: Current prices ;",
        "Gross domestic product: Current prices ;"
      ),
    series_type == "Seasonally Adjusted"
  ) |>
  pivot_wider(
    id_cols = date,
    names_from = series,
    values_from = value,
    names_repair = janitor::make_clean_names
  ) |>
  mutate(
    population = gross_domestic_product_current_prices /
      gdp_per_capita_current_prices,
    gross_disposable_income_per_capita = gross_disposable_income / population,
    gross_disposable_income_per_capita_annual = slider::slide_dbl(
      gross_disposable_income_per_capita,
      ~ sum(.x),
      .before = 3
    )
  ) |>
  inner_join(hpi) |>
  left_join(mean_house_price) |>
  mutate(
    mean_house_price = ifelse(
      is.na(mean_house_price),
      yes = mean_house_price[date == "2011-09-01"] *
        house_price_index /
        house_price_index[date == "2011-09-01"],
      no = mean_house_price
    ),
    ratio = 1000 * mean_house_price / gross_disposable_income_per_capita_annual
  )

# Required data for house finance stuff:
lending <- read_abs("5601.0", tables = 13, path = "data") |>
  separate_series(
    column_names = c(
      "households",
      "housing_finance",
      "investor",
      "purpose",
      "new_loan_commitments",
      "unit"
    ),
    remove_nas = T
  )

fhb_share <- read_abs(cat_no = "5601.0", tables = 1) |>
  filter(
    series_type == "Seasonally Adjusted",
    series %in%
      c(
        "Households ;  Housing Finance ;  Owner occupier ;  Total dwellings excluding refinancing ;  New loan commitments ;  Number ;",
        "Households ;  Housing Finance ;  Investor ;  Total dwellings excluding refinancing ;  New loan commitments ;  Number ;"
      )
  )
```

::: notes
We're going to talk about the housing market, house prices, and the changes made in relation to housing in the 2026 Budget
:::

## House Prices and Affordability

::: notes
House prices have grown much faster than incomes in Australia. The median household income per year (before-tax) in September 2025 was estimated at about \$105,000. Meanwhile, the median dwelling value was \$860,000 and the median weekly rent was \$670.

The average time required to save up a 20% deposit for the average dwelling is 11 years. While saving for the deposit, you're committing 33% of your income on rent. After securing a loan, you're committing 45% of your income just to service the mortgage.
:::

- It is more difficult than ever for first home buyers.

- Household income has not grown as fast as house prices

- Saving for a deposit is also more difficult

## House Prices and Affordability - Historical

::: notes
How has the picture changed over time?
:::

```{r}
gdi_pc |>
  filter(!is.na(ratio)) |>
  ggplot(aes(x = date, y = ratio)) +
  geom_line() +
  scale_x_date(breaks = plot_date_range(gdi_pc), date_labels = "%b-%Y") +
  theme_nrc() +
  labs(
    x = NULL,
    y = NULL,
    caption = "Source: ABS 5206.0 (Tables 1, 2, 20), ABS 6432.0 (Table 1), BIS (Q.AU.N.628)",
    title = "Ratio of Average House Price to Household Diposable Income"
  )
```