---
title: "Inflation and the Cost of Living"
format: html
fig-width: 6
fig-asp: 0.618
fig-format: retina
execute: 
  echo: false
  warning: false
editor: visual
---

```{r}
#| label: setup

library(tidyverse)
library(readabs)
library(reportabs)
library(directlabels)
library(scales)
library(glue)
library(gt)
library(geomtextpath)

cpi_quarterly_group <- read_absdata("cpi_quarterly_group", export_dir = "data", refresh = TRUE)

wpi_quarterly <- read_absdata("wpi_quarterly", export_dir = "data", refresh = TRUE)

cli_quarterly <- read_absdata("cli", export_dir = "data", refresh = TRUE)

plot_real_inflation_group <- function(cpi_category, since) {
  
  if (cpi_category == "Housing") {
    cpi <- cli_quarterly |> 
      filter(cpi_category == "Housing",
             data_type == "Index Numbers",
             household_type == "Employee households",
             date >= {{since}}) |> 
      mutate(value = value/value[date=={{since}}] - 1,
             indicator = "Cost of Living Index") |> 
      select(date, indicator, value)
  } else {
  cpi <- cpi_quarterly_group |> 
    filter(cpi_category == {{cpi_category}},
           unit == "Index Numbers",
           date >= {{since}}) |> 
    mutate(value = value/value[date=={{since}}] - 1,
           indicator = "Consumer Price Index") |> 
    select(date, indicator, value)
  }
  
  wpi <- wpi_quarterly |> 
    filter(data_type == "Quarterly Index", 
           date >= {{since}},
           state == "Australia",
           indicator == "Ordinary time hourly rates of pay excluding bonuses",
           sector == "Private and Public",
           industry == "Manufacturing") |> 
    mutate(value = value/value[date=={{since}}] -1,
           indicator = "Manufacturing Wages") |> 
    select(date, indicator, value)
  
  date_label <- label_date("%B %Y")(last(cpi$date))
  since_label <- label_date("%B %Y")(as_date(since))
  cpi_label <- label_percent(accuracy = 0.1)(last(cpi$value))
  wpi_label <- label_percent(accuracy = 0.1)(last(wpi$value))
  
  plot_data <- bind_rows(cpi, wpi)  |>
    group_by(indicator) |> 
    mutate(label = label_percent(accuracy = 0.1)(last(value))) |> 
    ungroup()
  
  inc_dec <- ifelse(last(cpi$value) > 0, "increased", "decreased")
  
  ggplot(plot_data, aes(x = date, y = value, col = indicator)) + 
    geom_line() +
    labs(x = NULL,
         y = NULL,
         title = glue("{cpi_category} {inc_dec} by {cpi_label} between {since_label} and {date_label}"),
         subtitle = glue("Manufacturing wages increased by {wpi_label} between {since_label} and {date_label}")) + 
    scale_x_date(breaks = plot_date_range(plot_data),
                 date_labels = "%b-%y") +
    scale_y_continuous(labels = label_percent()) +
    geom_dl(aes(label = label), method = list(cex = 10, "last.qp")) +
    theme_nrc(legend = "bottom")
  
  
}

cost_of_living_table <- function(since) {
  
  wage_growth <- wpi_quarterly |> 
    filter(date %in% c(max(date), {{since}}),
           data_type == "Quarterly Index",
           indicator == "Ordinary time hourly rates of pay excluding bonuses",
           state == "Australia",
           industry == "Manufacturing",
           sector == "Private and Public") |> 
    mutate(pc = value / value[date=={{since}}] - 1) |> 
    filter(date == max(date)) 
  
  first_date <- label_date(format = "%B %Y")(as_date({{since}}))
  last_date <-  label_date(format = "%B %Y")(wage_growth$date)
  
  cli_quarterly |> 
    filter(data_type == "Index Numbers",
           date %in% c(max(date), {{since}}),
           household_type == "Employee households") |> 
    group_by(cpi_category) |> 
    mutate(pc = value / value[date=={{since}}] - 1) |> 
    ungroup() |> 
    filter(date == max(date)) |> 
    select(cpi_category, pc) |> 
    add_row(cpi_category = "Wages",
            pc = wage_growth$pc) |> 
    arrange(-pc) |> 
    mutate(cpi_category = ifelse(cpi_category == "All groups", 
                                 "Average",
                                 cpi_category
                                 ),
           cpi_category = fct_inorder(cpi_category),
           cpi_category = fct_relevel(cpi_category, 
                                      c("Average"), 
                                      after = Inf
                                      )
           ) |> 
    arrange(cpi_category) |> 
    gt(rowname_col = "cpi_category") |> 
    tab_header(
      title = md("**Change in living costs by product category**"),
      subtitle = glue("% change between {last_date} and {first_date}")
    ) |> 
    tab_stubhead(label = "Price Category") |> 
    tab_style(style = cell_text(weight = "bold"),
              locations = cells_body(
                rows = cpi_category %in% c("Average", "Wages")
              )) |>  
    tab_style(style = cell_text(weight = "bold"),
              locations = cells_stub(
                rows = cpi_category %in% c("Average", "Wages")
              )) |> 
    tab_style(style = cell_text(align = "left"),
              locations = cells_stub()) |> 
    tab_style(style = cell_fill(color = "#FF7D38"),
              locations = cells_body(rows = cpi_category %in% c("Food and non-alcoholic beverages", "Housing", "Health"))) |> 
    fmt_percent(columns = pc, scale = 100, decimal = 1) |> 
    cols_label(pc = "% change")
  
}

```

# What is Inflation?

Inflation (officially the **Consumer Price Index**) measures changes in the price of a 'basket' of goods and services which account for a high proportion of expenditure by metropolitan households.

```{r}
#| label: fig-cpi-wpi
#| fig-cap: Inflation and Wages, March 2020 - December 2025


plot_real_inflation_group("All groups CPI", "2020-03-01")

```

# Inflation and the Cost of Living

Inflation is designed to measure price changes for the household sector as a whole. It is not an accurate reflection of how much your disposable income can buy. A more accurate measure is the **Living Cost Index.** This measures *by how much your disposable income would have to change to afford the same goods today as in the past*.

```{r}

cost_of_living_table(since = "2020-03-01") |> 
  tab_footnote(locations = cells_body(columns = pc, rows = 4),
               placement = "right",
               "To afford the same amount of food in December 2025 as you did in March 2020, your pay would have had to increase by 24%")

```

# Breaking News

While the Consumer Price Index is not useful for understanding changes to the cost of living, it is updated more frequently, andcan provide information about future cost of living increases.

However this will not capture any changes in prices resulting from the closure of the Strait of Hormuz.

```{r}
#| label: fig-monthly-cpi
#| fig-cap: Price Inflation, Selected Goods and Services, Last 12 months
#| fig-width: 10

monthly_cpi <- read_abs("6401.0", tables = "2")

monthly_cpi_plot <- monthly_cpi  |> 
  separate_series() |> 
  filter(series_1 == "Percentage Change from Corresponding Month of Previous Year",
         series_3 == "Australia",
         date == max(date)) |>
  arrange(value) |> 
  mutate(series_2 = fct_inorder(series_2),
         nudge = max(value)*0.075 + max(value) - value)

monthly_cpi_plot |> 
  ggplot(aes(x = value, y = series_2, label = scales::label_percent(scale = 1)(value))) + 
  geom_col() +
  geom_text(nudge_x = monthly_cpi_plot$nudge) + 
  theme_nrc(flipped = TRUE) +
  labs(x = "% change from February 2025",
       y = NULL) +
  scale_x_continuous(labels = scales::label_percent(scale = 1)) 

```
